---
title: "GBM_report_AhmedRayan_C2168667"
author: 'Rayan Ahmed '
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Libaray Required}
library(Seurat)
library(patchwork)
library(dplyr)
```

# Seurat Object

### Load the Dataset into Seurat Object

```{r Load the Dataset into Seurat Object }
main <- read.table('Data/GSE84465_GBM_All_data.csv', sep=" ", header=TRUE)
object_seurat <- CreateSeuratObject(counts = main, project = "GSE84465")
object_seurat
slotNames(object_seurat)
```

```{r }
object_seurat[["percent.mt"]] <- PercentageFeatureSet(object_seurat, pattern = "^MT-")

```

### Dataset Quality Check

```{r Data Quality Check}
# Visualize QC metrics as a violin plot
VlnPlot(object = object_seurat, features = c("nFeature_RNA", "nCount_RNA", "percent.mt", ncol = 3))
```

### Feature Quality Check

```{r Feature Quality Check }
# Feature scatter plot
plot1 <- FeatureScatter(object_seurat, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(object_seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```

### Normalization of Data

```{r Normalizing Data}
object_seurat <- NormalizeData(object_seurat, normalization.method = "LogNormalize", scale.factor = 10000)
```

### Identification of highly variable Genes

```{r Identify HVGs}
# Identification of highly variable features (feature selection)
object_seurat <- FindVariableFeatures(object_seurat, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(object_seurat), 10)
#write.csv(top10, file="top_hgv.csv", row.names=FALSE)
# plot variable features with and without labels
plot1 <- VariableFeaturePlot(object_seurat)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2
```

### Scale The Data

```{r Scaling Data}
all.genes <- rownames(object_seurat)
object_seurat <- ScaleData(object_seurat, features=all.genes)
```

### Perform linear dimensional reduction

```{r Dimensional reduction Using PCA}
# linear dimensional reduction
object_seurat <- RunPCA(object_seurat, features = VariableFeatures(object = object_seurat))
```

### Plot Heatmap For PCs

```{r Heatmap for top 15 PCs}
DimHeatmap(object_seurat, dims = 1:15, cells = 500, balanced = TRUE)
```

### Plot For PCs

```{r PCA Plot}
DimPlot(object_seurat, reduction = "pca")
```

### Statistical Significance of PCA scores

```{r statistical significance of PCA scores}
# Dimensionality of the dataset
object_seurat <- JackStraw(object_seurat, num.replicate = 100)
object_seurat <- ScoreJackStraw(object_seurat, dims = 1:20)
JackStrawPlot(object_seurat, dims = 1:15)
```

### Cluster Cells

```{r Cluster Cells}
#Cluster cells
object_seurat <- FindNeighbors(object_seurat, dims= 1:10)
object_seurat <- FindClusters(object_seurat, resolution = 0.5)
head(Idents(object_seurat), 10)
```

### Dimensionality reduction - UMAP Method

```{r dimensionality reduction UMAP }
# Run non-linear dimensional reduction (UMAP)
object_seurat <- RunUMAP(object_seurat, dims = 1:10)
DimPlot(object_seurat, reduction = "umap")
```

### Total Gene Across all Clusters Histogram

```{r Gene count across all Clusters}
# Get the cluster labels
cluster_labels <- Idents(object_seurat)

# Count the number of cells in each cluster
cluster_counts <- table(cluster_labels)

# Create a data frame for ggplot
cluster_data <- data.frame(Cluster = names(cluster_counts), Count = as.numeric(cluster_counts))

# Create the histogram
library(ggplot2)
ggplot(cluster_data, aes(x = Count, y = Cluster)) +
  geom_col() +
  xlab("Number of Cells") +
  ylab("Cluster") +
  theme_minimal()

print(cluster_data)
#write.csv(cluster_data, file="cluster_data.csv", row.names=FALSE)
```

### Markers distinguishing in Clusters

```{r Finding Markers in Clusters}
# find all markers distinguishing cluster 0 from cluster 2
markers_for_cluster0 <- FindMarkers(object_seurat, ident.1 = 0, ident.2 = 2, min.pct = 0.25)
head(markers_for_cluster0, n=5)
```

### Naming the Clusters

```{r Lableling the Clusters}
updated_cluster_ids <- c("Myeloid", "Cancer", "Macrophages", "Unknown", "Myeloid", "Cancer", "Microglia", "Microglia", "Cancer", "Unknown", "Unknown", "Unknown", "Oligodendrocyte", "Unknown", "Astrocyte", "Unknown", "Cancer", "Neurons")
names(updated_cluster_ids) <- levels(object_seurat)
object_seurat <- RenameIdents(object_seurat, updated_cluster_ids)
DimPlot(object_seurat, reduction = "umap", label = TRUE, pt.size = 0.5) + NoLegend()
```

### Differential Expression for Features

```{r Differential Expression for Features}
# Find differentially expressed features (cluster biomarkers)
seurat_markers <- FindAllMarkers(object_seurat, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
seurat_markers %>% group_by(cluster) %>% top_n(n=5, wt=avg_log2FC)
top_markers<- seurat_markers %>% group_by(cluster) %>% top_n(n=3, wt=avg_log2FC)
#write.csv(top_markers, file="top_markers.csv", row.names=FALSE)
```

### Heatmap for Top 3 Marker in Clusters

```{r Heatmap for Top 3 Marker accross Clusters}
top3 <- seurat_markers %>% group_by(cluster) %>% top_n(n=3, wt=avg_log2FC)
DoHeatmap(object_seurat, features = top3$gene) 
```

### Differential Expression Using seurat

```{r Differential Expression Using seurat}
# Find differentially expressed features between Myeloid and  Macrophages
myeloid_markers <- FindMarkers(object_seurat, ident.1 = "Myeloid", ident.2 = "Macrophages")
# view results
head(myeloid_markers, 10)
```

### Volcano Plot For seaurt identified DEGs

```{r Volcano Plot For seaurt DEGs}
#volcano plot for myeloid_markers
#install.packages("melt")
#library(melt)
# Convert the data frame to a long format for plotting
#myeloid_markers_long <- melt(myeloid_markers, id.vars = c( "p_val", "avg_log2FC", "pct.1", "pct.2", "p_val_adj"))

# Create a volcano plot
#ggplot(myeloid_markers_long, aes(x = avg_log2FC, y = -log10(p_val), color = ifelse(abs(avg_log2FC) > 1, "Significant", "Not Significant"))) +
 # geom_point() +
 # theme_classic() +
  #labs(x = "Log2 Fold Change", y = "-Log10 P-Value", color = "Significance") +
  #scale_color_manual(values = c("blue", "red")) +
  #theme(legend.position = "bottom")

```

### Differential Expression Using DESeq2

```{r Differential Expression Using DESeq2}
### Differential expression testing using DESeq2
markers_degs <- FindMarkers(object_seurat, ident.1 = "Myeloid", ident.2 = "Macrophages", test.use = "DESeq2", max.cells.per.ident = 50)

# view results
head(markers_degs, 10)
                   
```

### Volcano Plot For DESeq2 identified DEGs

```{r Volcano Plot For DESeq2 DEGs}
#volcano plot for markers_degs
# Convert the data frame to a long format for plotting
#markers_degs_long <- melt(markers_degs, id.vars = c( "p_val", "avg_log2FC", "pct.1", "pct.2", "p_val_adj"))

# Create a volcano plot
#ggplot(markers_degs_long, aes(x = avg_log2FC, y = -log10(p_val), color = ifelse(abs(avg_log2FC) > 1, "Significant", "Not Significant"))) +
  #geom_point() +
  #theme_classic() +
  #labs(x = "Log2 Fold Change", y = "-Log10 P-Value", color = "Significance") +
  #scale_color_manual(values = c("blue", "red")) +
  #theme(legend.position = "bottom")
```

### Dot Plot for DEGs in Myeliod and Macrophages

```{r Dot Plot for DEGs in My and Marco}

my_and_mac <- c('NFKBID','KIAA1274','GJA1','C3AR1','LYZ','IRF8','HSPA7','IGFBP2','SIGLEC10','CA2')

DotPlot(object_seurat, features = my_and_mac , idents = c("Myeloid", "Macrophages")) + RotatedAxis()
```

### Heatmap for DEGs across all clusters

```{r Heatmap for DEGs against all Clusters}
### Heatmap for differentially expressed genes among all genes
DoHeatmap(subset(object_seurat, downsample =100), features = my_and_mac, size = 4)
```

### Violin plot for DEGs across all clusters

```{r Violin Plot for DEGs against all clusters}
### Violin plot for differentially genes among all cell groups
for (i in 1:10){
  print(VlnPlot(object_seurat, features = seurat_markers$gene[i]), pt.size=0.1) + NoLegend()
}
```

### Dot Plot for DEGs across all clusters

```{r Dot Plot for DEGs against all clusters}
### Dot plot for differentially genes among all cell groups
for (i in 1:10){
  print(DotPlot(object_seurat, features = seurat_markers$gene[i])) + RotatedAxis()
}
```
